The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-02-04 Paper.Rmd
changed the description of what was done with the first set of models in the Q2 section

<!--***DAN: made it to here, 2018 1 10, 1252pm-->

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explore via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the idea that non-N copula structure and asymmetric tail dependence 
between environmental variables measured in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables. Let $P_i(t)$ be a population index for location $i$ ($i=1,2$),
and assume the index follows AR(1) dynamics within a patch, 
$P_i(t+1)=\beta P_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$. Here the $\epsilon_i(t)$ are 
standard-normal random variables that are independent for distinct times, $t$, but 
that are non-independent across patches.

<!--***DAN: Note to self, this model, and also the other one we study in this section,
are actually much more general than populatin. Should replace P_i with E_i where E_i is
just some ecological variable. Could be deviaitons from a pop from carrying capacity, could be a 
biogeochemical variable, could maybe even be an evolutinary variable. Anything that is
affected by an environmental variable and could reasonably be modelled by an AR process. 
So the generalization to biogeochemical and community
data seems pretty straightforward. Need to think this through and reword the corresponding
part of the intro to encompass this greater generality. How would the evolutionary case work?
Well, suppose the two "patches" are actually traits, and the perturbations epsilon are somehow
the "drift events" or something like that? This does not really make sense, strictly speaking,
now that I am thinking about it, but maybe the model could be modified to make it make sense?
At least I have straightforward generalization to biogeochem and community cases. 
More on evolutionary: suppose we model traits as a bivariate random walk where the perturbations
in the traits are related with a copula with some tail dependence. Suppose you have a tree and 
you model evolution down the tree in this way and then look at the bivariate distribution of
trait on the tips of the tree. What do you get? I guess you get some copula structure?-->

To begin with, we chose a two habitat patch ($i = 1, 2$) with AR(1) model [Eq. \ref{eq.AR1}] where population $P_{i}$ at time $(t+1)$ depends upon its past value at $(t)$ and the iid noise (assumed to be caused by environmental fluctuations) $\epsilon_{i} (t)$.

\begin{equation}\label{eq.AR1}
P_{i}(t+1)=\beta P_{i}(t)+\alpha \epsilon_{i}(t)
\end{equation}

To keep $Var(P_{i}(t+1))$ equal to unity, we set noise coefficient $\alpha$ in terms of AR coefficient $\beta$ as $\alpha = \sqrt(1-\beta^2)$. We showed with Eq. \ref{eq.AR1} in the next section that both model selection method and non-parametric analysis indicated similar tail dependence in population copula as that of noise copula. <!--Asymmetric sensitivity and copula structure in populations-->
\textbf{(ii)}

>>>>

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we explored via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables.

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t-1)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable.

To explore the importance of nonlinear dynamics for our hypothersis, we also considered 
nonlinear population models based on $P_i(t+1)=P_i(t)\exp \left(r\left(1-P_i(t)/K  \right) \right)+\sigma\epsilon_i(t))$, using $r=0.5$, $K=100$, and $\sigma=0.1$ or 
$\sigma=1$. When $\sigma=0.1$, population dynamics stay close to the carrying capacity, 
$K$, and the nonlinearities of the model have limited influence on dynamics. When
$\sigma=1$, model dynamics are strongly nonlinear. We refer to these as
the weak-noise and strong-noise cases. The $epsilon_i(t)$ were again standard
normal and independent through time, but were related (see below) across locations.

For each of the three model setups, for each of the copulas C, SC, G, SG, J, SJ,
and for each Kendall correlation value $\tau=0.1,0.2,\ldots,0.9$, 
we generated 5000 noise pairs $(\epsilon_1(t),\epsilon_2(t))$ from the
bivariate random variable with the given copula family and the given $\tau$.
We then used the generated noise to drive the model, and retained both the
noise and population values for the final 500 
time steps of the simulation. For each model, copula family, and $\tau$, the following
statistics were computed for both the noises and populations: 
Pearson, Spearman and Kendall correlations, $\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, 
$\Dsq_l$, $\Dsq_u$, $\cor_l - \cor_u$, $\Ps_l - \Ps_u$, and $\Dsq_u - \Dsq_l$. 
Values were plotted for noises and populations against $\tau$. If the hypothesis is
correct that copula structure between locations of environmental noise can be 
transmitted to produce similar dependence structure for ecological variables, 
plots should be similar for populations and noises.

<!--***DAN: Note to self, this model, and also the other one we study in this section,
are actually much more general than populatin. Should replace P_i with E_i where E_i is
just some ecological variable. Could be deviaitons from a pop from carrying capacity, could be a 
biogeochemical variable, could maybe even be an evolutinary variable. Anything that is
affected by an environmental variable and could reasonably be modelled by an AR process. 
So the generalization to biogeochemical and community
data seems pretty straightforward. Need to think this through and reword the corresponding
part of the intro to encompass this greater generality. How would the evolutionary case work?
Well, suppose the two "patches" are actually traits, and the perturbations epsilon are somehow
the "drift events" or something like that? This does not really make sense, strictly speaking,
now that I am thinking about it, but maybe the model could be modified to make it make sense?
At least I have straightforward generalization to biogeochem and community cases. 
More on evolutionary: suppose we model traits as a bivariate random walk where the perturbations
in the traits are related with a copula with some tail dependence. Suppose you have a tree and 
you model evolution down the tree in this way and then look at the bivariate distribution of
trait on the tips of the tree. What do you get? I guess you get some copula structure? maybe
this thinking can lead to some isnights into how copula structure we observed in the species
trait data might have come about?-->
